Circumnuclear Media and Accretion Rates of Quiescent Supermassive Black Holes

### 2015-04-05 ms.tex
salvaged appendix B
intro again

Understanding why most SMBHs appear to be inactive requires
characterizing their gaseous environments.  Gas near the sphere of
influence of the SMBH, hereafter the `circumnuclear medium' (CNM),
controls the mass accretion rate, $\dot{M}$.  The accretion rate in
turn sets the luminosity of the black hole and the feedback of this
energy and momentum output on larger physical scales.
The presence of dense gas in the nuclear region may lead to runaway
cooling, resulting in bursty episodes of star formation and AGN
activity (e.g.~\citealt{Ciotti&Ostriker07}; \citealt{Ciotti+10}).

Knowledge of how $\dot{M}$ depends on the SMBH mass, $\Mbh$, and other
properties of the galactic nucleus informs key questions related to
the co-evolution of SMBHs and their host galaxies with cosmic time
(e.g.~\citealt{Kormendy&Ho13}; \citealt{HeckmanBest:2014a}).  SMBH growth
in the low redshift universe is dominated by low mass black holes,
$M_{\bullet} \lesssim 10^{8}M_{\odot}$ (e.g.~\citealt{Heckman+04}), a
fact which is often attributed to the general trend of `cosmic
down-sizing' resulting from hierarchical structure growth (e.g,
\citealt{Gallo+08}).  However, the physical processes involved in
growing low mass blacks may be distinct from those operating at higher
SMBH masses and accretion rates in quasars.  A key question is
whether SMBHs grow primarily by accreting gas fed in directly from
galactic or extra-galactic scales, or whether significant growth
results also from internal, secular processes such as local stellar
mass loss in the nuclear region.

A better understanding of what mechanisms regulate accretion onto
quiescent SMBHs would have implications for a variety of observations,
such as constraints on the occupation fraction of SMBHs in low mass
galaxies.  \citet{Miller+15} use the average relationship between the
nuclear X-ray luminosities, $L_{X}$, of a sample of elliptical
galaxies and their associated SMBH masses to infer tentative evidence for a
SMBH occupation fraction less than unity in galaxies with stellar
masses $M_{\star} \lesssim 10^{10}M_{\odot}$ ($M_{\bullet} \lesssim
10^{7}M_{\odot}$).  This method relies on extrapolating a power-law
fit of the $L_X-\Mbh$ distribution to low values of $L_{\rm X}$ below
the instrument detection threshold, an assumption which is questionable if
different physical processes control the accretion rates onto the
lowest mass SMBHs.

@@

\citet{Ho:2009a} determines the SMBH accretion rates in a sample of
early-type galaxies using X-ray observations to determine the
Bondi accretion rate and based on estimates for the mass loss rate of
evolved stars.  Both methods lead to the conclusion that the available
gas reservoir is more than sufficient to power the observed
low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative
efficiency for thin disk accretion.  Several lines of evidence now suggest that low-luminosity AGN result from accretion
proceeding in a radiatively inefficient mode
(\citealt{Yuan&Narayan14}), due either to the advection of
gravitationally-released energy across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to the low efficiency with which the inflowing gas ultimately reaches the SMBH due to disk outflows (e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Past studies, employing multi-dimensional numerical hydrodynamics and
including variety of (parameterized) physical effects, have focused on
massive elliptical galaxies (e.g.~\citealt{Ciotti&Ostriker07};
\citealt{Ciotti+10}).  A key finding of these works is the periodic
development of cooling instabilities on galactic scales, which
temporarily increase the gas accretion rate towards the nucleus until
feedback becomes effective to shut off the flow and halt
SMBH growth.  Here we instead focus on time-independent models,
neglecting radiative cooling.  This approach allows us to
systematically explore the relevant parameter space and to derive
simple analytic expressions that prove useful to determine whether
cooling instabilities manifest across the expected range of galaxy
properties, or whether other (non-AGN) forms of feedback can result in
steady accretion.  Even if cooling instabilities develop on galactic
scales over longer $\sim$ Gyr timescales, a state of
quasi-steady accretion may exist between these inflow events on
smaller radial scales comparable to the sphere of influence.  We are
thus motivated to explore whether steady flow is typically expected within the
nucleus under, or whether in most galaxies the nuclear gas content is always
building towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

@@

This paper is organized as followed.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxy properties used in our
analysis ($\S\ref{sec:gal_model}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe some
analytic results, while in $\S\ref{sec:numerical}$ we describe our
numerical results.  In $\S\ref{sec:discussion}$ we discuss the
implications of our results, to topics including the nuclear X-ray
luminosities of quiescent black holes, jetted TDEs, and the growth
time of SMBHs.  In $\S\ref{sec:conclusions}$ we summarize our
conclusions.  Table \ref{table:definitions} provides the definitions
of commonly used variables.  Appendices \ref{app:rs} and \ref{app:be}
provide useful analytic results for the stagnation radius and the
Bernoulli parameter of the flow, while Appendix
\ref{app:windheat} describes our method for calculating stellar wind
heating and mass input in greater detail.

@@

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random
movement of stars with respect to one another at the characteristic
velocity of 
\be\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2},
\ee
where the first term is the velocity dispersion of the power law stellar density profile in the SMBH potential.  The second term, $\sigma_0\approx 3 \sigma^2$ is a constant, motivated by the observed velocity dispersion outside of the SMBH sphere of influence \citep{ChaeBernardi+:2014a}, where $\sigma$ is the velocity dispersion in the $\Mbh-\sigma$
relation (eq.~[\ref{eq:Msigma}]).  The second component of the total heating, $v_{w}^{2}$, parameterizes all additional sources of energy input, including faster winds from young stars, millisecond pulsars, supernovae, AGN feedback, etc.  We assume that $v_w$ is constant with radius, i.e. that the volumetric heating rate is proportional to the local stellar density.  The value of $v_w$ will in general depend on the SMBH mass and the age of the stellar population ($\S\ref{sec:heating}$).

@@

  \clearpage
  \appendix
  \section{Analytic expression for the stagnation radius}
  \label{app:rs}
  \input{rs}

  \section{Analytic expression for Bernoulli parameter of gas}
  \label{app:be}
  \input{be}

>>>>

Understanding why most SMBHs appear to be inactive requires
characterizing their gaseous environments.  Gas near the SMBH sphere of
influence, hereafter denoted the `circumnuclear medium' (CNM),
controls the mass accretion rate, $\dot{M}$.  The accretion rate in
turn determines the SMBH luminosity and the feedback of its energy and momentum output on larger scales.
The presence of dense gas in the nucleus may lead to runaway cooling, resulting in bursty episodes of star formation and AGN
activity (e.g.~\citealt{Ciotti&Ostriker07}; \citealt{Ciotti+10}).

Knowledge of how $\dot{M}$ depends on the SMBH mass, $\Mbh$, and other
properties of the nucleus informs key questions related to
the co-evolution of SMBHs and their host galaxies with cosmic time
(e.g.~\citealt{Kormendy&Ho13}; \citealt{HeckmanBest:2014a}).  In the low redshift universe, SMBH growth
is dominated by low mass black holes,
$M_{\bullet} \lesssim 10^{8}M_{\odot}$ (e.g.~\citealt{Heckman+04}), a
fact often attributed to the trend of `cosmic
down-sizing' resulting from hierarchical structure growth (e.g,
\citealt{Gallo+08}).  However, the physical processes by which low mass black holes accrete could in principle be distinct from those operating at higher SMBH masses and accretion rates in high redshift quasars.  One key question is
whether SMBHs grow primarily by the accretion of gas fed in directly from
galactic or extra-galactic scales, or whether significant growth
can result also from local stellar mass loss in the nuclear region.

A better understanding of what mechanisms regulate accretion onto
quiescent SMBHs would shed new light on a variety of observations,
such as the occupation fraction of SMBHs in low mass
galaxies.  \citet{Miller+15} use the average relationship between the
nuclear X-ray luminosities, $L_{X}$, of a sample of early type
galaxies and their associated SMBH masses to infer tentative evidence that the
SMBH occupation fraction becomes less than unity for galaxies with stellar
masses $M_{\star} \lesssim 10^{10}M_{\odot}$ ($M_{\bullet} \lesssim
10^{7}M_{\odot}$).  This method relies on extrapolating a power-law
fit of the $L_X-\Mbh$ distribution to low values of $L_{\rm X}$ below
the instrument detection threshold, an assumption which is questionable if
different physical processes control the accretion rates onto the
lowest mass SMBHs.

@@

\citet{Ho:2009a} determines the SMBH accretion rates in a sample of
early-type galaxies using X-ray observations to determine the
Bondi accretion rate and based on estimates for the mass loss rate of
evolved stars.  Both methods lead him to conclude that the available
gas reservoir is more than sufficient to power the observed
low-luminosity AGN, assuming the standard $\sim$ 10 per cent radiative
efficiency for thin disk accretion.  Several lines of evidence now suggest that low-luminosity AGN result from accretion
proceeding in a radiatively inefficient mode
(\citealt{Yuan&Narayan14}), due either to the advection of
gravitationally-released energy across the SMBH horizon
(e.g.~\citealt{Narayan&Yi95}) or due to the low efficiency with which the inflowing gas ultimately reaches the SMBH due to disk outflows (e.g.~\citealt{Blandford&Begelman99}; \citealt{Li+13}).

@@

Previous studies, employing multi-dimensional numerical hydrodynamics and
including variety of (parameterized) physical effects, have focused on
massive elliptical galaxies (e.g.~\citealt{Ciotti&Ostriker07};
\citealt{Ciotti+10}).  These works show the periodic
development of cooling instabilities on galactic scales, which
temporarily increase the gas accretion rate towards the nucleus until
feedback becomes effective to shut off the flow and halt
SMBH growth.

In this paper we instead focus on time-independent models,
neglecting radiative cooling.  This approach allows us to
systematically explore the relevant parameter space and to derive
analytic expressions that prove useful in determining whether
cooling instabilities manifest across the expected range of galaxy
properties, or whether other (non-AGN) forms of feedback can produce a prolonged state of steady, thermally stable accretion.  Even if cooling instabilities develop on galactic scales over longer $\sim$ Gyr timescales, a quasi steady-state  may exist between these inflow events on
smaller radial scales comparable to the sphere of influence.  We are
thus motivated to explore whether such steady flows are generally expected within the
nucleus under, or whether in most galaxies the nuclear gas content is always
building towards instability.
%A chief goal of our analysis is to provide a framework for interpreting the properties of quiescent galactic nuclei given the comparatively well understood physics of stellar winds and supernova heating.

@@

This paper is organized as follows.  In $\S\ref{sec:model}$ we
describe our model, including the sample of galaxy properties used in our
analysis ($\S\ref{sec:gal_model}$) and our numerical procedure for
calculating the steady-state hydrodynamic profile of the CNM
($\S\ref{sec:hydro}$).  In $\S\ref{sec:results}$ we describe our analytic results, which are justified via numerical solutions in $\S\ref{sec:numerical}$ we describe our
numerical results.  In $\S\ref{sec:discussion}$ we discuss the
implications of our results for topics which include the nuclear X-ray
luminosities of quiescent black holes, jetted TDEs, and the growth of SMBHs in the local universe.  In $\S\ref{sec:conclusions}$ we summarize our conclusions.  Table \ref{table:definitions} provides the definitions
of commonly used variables.  Appendices \ref{app:rs} and \ref{app:be}
provide useful analytic results for the stagnation radius and the
Bernoulli parameter of the flow, while Appendix \ref{app:windheat} provides the details of our method for calculating stellar wind heating and mass input.

@@

The term $\propto \vw^2 = \sigma(r)^2+v_{w}^2$ in the entropy equation
accounts for external heating sources (e.g.,
\citealt{ShcherbakovWong+:2014a}).  The first term approximates the
minimum amount of shock heating from stellar winds due to the random
movement of stars with respect to one another at the characteristic
velocity of 
\be\sigma \approx \sqrt{\frac{3 G \Mbh}{(\Gamma+2)
    r}+\sigma_0^2},
\label{eq:sigmarel}
\ee
where the first term is the velocity dispersion of the power law stellar density profile in the SMBH potential.  The second term, $\sigma_0\approx 3 \sigma^2$ is a constant, motivated by the observed velocity dispersion outside of the SMBH sphere of influence \citep{ChaeBernardi+:2014a}, where $\sigma$ is the velocity dispersion in the $\Mbh-\sigma$
relation (eq.~[\ref{eq:Msigma}]).  The second component of the total heating, $v_{w}^{2}$, parameterizes all additional sources of energy input, including faster winds from young stars, millisecond pulsars, supernovae, AGN feedback, etc.  We assume that $v_w$ is constant with radius, i.e. that the volumetric heating rate is proportional to the local stellar density.  The value of $v_w$ will in general depend on the SMBH mass and the age of the stellar population ($\S\ref{sec:heating}$).

@@

  \clearpage
  \appendix
  \section{Derivation of stagnation radius}
  \label{app:rs}
  \input{rs}

  \section{Analytic expression for Bernoulli parameter}
  \label{app:be}
  \input{be}
